---
title: "22100 - R for Bio Data Science"
subtitle: "Lab 7 - Scripting in a Reproducible and Collaborative Framework using GitHub via RStudio: Assignment 5: PCA Micro Project"
author: "Leon Eyrich Jessen"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: images/r_for_bio_data_science_hex_logo_quadratic_small.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("broom")
library("ggrepel")
```

## Let us take a look at the BLOSUM62 matrix

```{r message=FALSE, warning=FALSE}
bl62 <- read.table(
  file = "https://www.ncbi.nlm.nih.gov/Class/FieldGuide/BLOSUM62.txt",
  comment = "#") %>%
  rownames_to_column(var = "aa") %>%
  as_tibble
bl62
```

## Wrangle and save the data

```{r}
std_aa = c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
bl62 <- bl62 %>%
  select(aa, all_of(std_aa)) %>%
  filter(aa %in% std_aa)
write_tsv(x = bl62,
          file = "data/BLOSUM62_ncbi.tsv")
```

## Wrangle and save the data
```{r}
bl62
```

## Create a PCA object

```{r}
bl62_pca <- bl62 %>%
  select(-aa) %>%
  prcomp(center = TRUE, scale = TRUE)
bl62_pca
```

## Make a scree plot using broom to tidy

<div style="float: left; width: 50%;">
```{r}
bl62_pca %>%
  tidy("pcs")
```
</div>

<div style="float: right; width: 50%;">
```{r fig.align="left", fig.width=5, fig.height=4}
bl62_pca %>%
  tidy("pcs") %>% 
  ggplot(aes(x = PC,
             y = percent)) +
  geom_col() +
  theme_bw()
```
</div>

## Augment using broom

```{r}
bl62_pca_aug <- bl62_pca %>%
  augment(bl62)
```

## Augment using broom
```{r echo=FALSE}
bl62_pca_aug
```

## Add some chemical classes

```{r}
get_chem_class <- function(x){
  chem_cols <- c("A" = "Hydrophobic", "R" = "Basic", "N" = "Neutral", "D" = "Acidic",
                 "C" = "sulphur", "Q" = "Neutral", "E" = "Acidic", "G" = "Polar",
                 "H" = "Basic", "I" = "Hydrophobic", "L" = "Hydrophobic", "K" = "Basic",
                 "M" = "sulphur", "F" = "Hydrophobic", "P" = "Hydrophobic", "S" = "Polar",
                 "T" = "Polar", "W" = "Hydrophobic", "Y" = "Polar", "V" = "Hydrophobic")
  return(factor(chem_cols[x])) # Custom function NOT a pipeline, hence the [] => shareable!
}
```

```{r}
bl62_pca_aug <- bl62_pca_aug %>% 
  mutate(chem_class = get_chem_class(aa))
```

#### Note how we are using our own custom function inside a dplyr pipeline here

```{r}
get_chem_class(x = c("A", "R", "D"))
```

## Add some chemical classes

```{r}
bl62_pca_aug %>% select(aa, chem_class)
```

## Plot the PCA Plot

```{r}
bl62_pca_aug %>% 
  ggplot(aes(x = .fittedPC1, y = .fittedPC2, label = aa, colour = chem_class)) +
  geom_text() +
  theme(legend.position = "bottom")
```

## Plot the Loadings Plot

<div style="float: left; width: 50%;">
```{r}
# define arrow style for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed",
  length = grid::unit(8, "pt"))

# plot rotation matrix
pl <- bl62_pca %>%
  tidy(matrix = "rotation") %>%
  pivot_wider(names_from = "PC",
              names_prefix = "PC",
              values_from = "value") %>%
  ggplot(aes(x = PC1,
             y = PC2,
             label = column)) +
  geom_segment(xend = 0,
               yend = 0,
               arrow = arrow_style) +
  geom_text(hjust = 1,
            nudge_x = -0.02, 
            color = "#904C2F") +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal(12)
```
</div>

<div style="float: right; width: 50%;">
```{r fig.align="left", fig.width=5, fig.height=4}
pl
```
</div>

## Plot the PCA

```{r}
bl62_pca_aug %>% pull(chem_class) %>% levels
```

```{r eval=FALSE}
# Bad code style to make it fit on slide
x <- bl62_pca %>% tidy("pcs") %>% filter(PC == 1) %>% pull(percent)
x <- str_c("PC1 (", round(x*100, 2), "%)")
y <- bl62_pca %>% tidy("pcs") %>% filter(PC == 2) %>% pull(percent)
y <- str_c("PC2 (", round(y*100, 2), "%)")
```


```{r eval=FALSE}
bl62_pca_aug %>% 
  ggplot(aes(x = .fittedPC1, y = .fittedPC2,
             label = aa, colour = chem_class)) +
  geom_label_repel() + # library("ggrepel") trick here
  theme(legend.position = "bottom") +
  scale_colour_manual(values = c("red", "blue", "black",
                                 "purple", "green", "yellow")) +
  labs(x = x, y = y)
```

## Plot the PCA

<div style="float: left; width: 50%;">
```{r echo=FALSE}
x <- bl62_pca %>% tidy("pcs") %>% filter(PC==1) %>% pull(percent)
x <- str_c("PC1 (", round(x*100, 2), "%)")
y <- bl62_pca %>% tidy("pcs") %>% filter(PC==2) %>% pull(percent)
y <- str_c("PC2 (", round(y*100, 2), "%)")
```

```{r echo=FALSE, fig.align="center", fig.width=4.7}
bl62_pca_aug %>% 
  ggplot(aes(x = .fittedPC1, y = .fittedPC2,
             label = aa, colour = chem_class)) +
  geom_label_repel() + # library("ggrepel") trick here
  theme(legend.position = "bottom") +
  scale_colour_manual(values = c("red", "blue", "black",
                                 "purple", "green", "yellow")) +
  labs(x = x, y = y, colour = "")
```
</div>
<div style="float: right; width: 50%;">
```{r echo=FALSE, fig.align="center", fig.width=4.7}
bl62_pca %>%
  tidy("pcs") %>% 
  ggplot(aes(x = PC, y = percent)) +
  geom_col() +
  theme_bw()
```
</div>

- Up: Aromatic, down: Aliphatic, Left: Charged, Right: Hydrophobic
- Non of this was given to the model, it is all in the BLOSUM62 numbers

## Add k-means

```{r}
bl62 %>%
  select(-aa) %>%
  kmeans(centers = 6, iter.max = 1000, nstart = 10) %>%
  augment(bl62_pca_aug) %>% 
  head
```

## Add k-means

<div style="float: left; width: 50%;">
```{r echo=FALSE, fig.align="center", fig.width=4.7}
bl62_pca_aug %>% 
  ggplot(aes(x = .fittedPC1, y = .fittedPC2,
             label = aa, colour = chem_class)) +
  geom_label_repel() + # library("ggrepel") trick here
  theme(legend.position = "bottom") +
  scale_colour_manual(values = c("red", "blue", "black",
                                 "purple", "green", "yellow")) +
  labs(x = x, y = y, colour = "")
```
</div>

<div style="float: right; width: 50%;">
```{r echo=FALSE, fig.align="center", fig.width=4.7}
set.seed(174369)
bl62 %>%
  select(-aa) %>%
  kmeans(centers = 6, iter.max = 1000, nstart = 10) %>%
  augment(bl62_pca_aug) %>% 
  ggplot(aes(x = .fittedPC1, y = .fittedPC2,
             label = aa, colour = .cluster)) +
  geom_label_repel() + # library("ggrepel") trick here
  theme(legend.position = "bottom") +
  scale_colour_manual(values = c("red", "blue", "black",
                                 "purple", "green", "yellow")) +
  labs(x = x, y = y, colour = "")

```
</div>

# Questions?


