---
title: "Midway Evalutation"
author: "Leon Eyrich Jessen"
format:
  revealjs:
    embed-resources: true
    theme: moon
    slide-number: c/t
    width: 1600
    height: 900
    mainfont: avenir
    logo: images/r4bds_logo_small.png
    footer: "R for Bio Data Science"
---

<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->


```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

library("tidyverse")
this_year <- "data/2025 Course Midway Evaluation (Responses) - Form Responses 1.tsv"
eval_data <- read_tsv(file = this_year)
mk_bar_plot <- function(data, x){
  m <- data |>
    summarise(m = median({{x}})) |>
    pull() 
  data |>
  count({{x}}) |> 
  mutate(pct = str_c(round(n/sum(n)*100,1), "%")) |> 
  ggplot(
    mapping = aes(
      x = {{x}},
      y = n,
      label = pct)) +
  geom_col(fill = "cornflowerblue",
           colour = "black") +
  geom_text(size = 6,
            nudge_y = 1) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(breaks = seq(1,10),
                     limits = c(0,11)) +
  labs(x = "",
       y = "",
       title = rlang::englue("{{x}}"),
       subtitle = str_c("Median Score = ", m)) +
  theme_minimal(base_size = 20) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(size = 14),
        plot.title.position = "plot",
        plot.title.align = 0)
}
```

## Midway Evaluation

- Our opportunity to understand if the course elements works as intended
- Your opportunity to let us know, how the course looks on the students side
- Super important to align
- 67 out of 140 enrolled students answered this year 
- THANK YOU for this important contribution!



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Highlights: What's Working Well in the Course

- Hands-on exercises and practical sessions are consistently praised as engaging, well-designed, and central to learning success.
- Instructor and TA support receive strong positive mentions—students find the guidance clear, helpful, and motivating.
- Collaborative learning through group work is viewed as effective for understanding material and maintaining engagement.



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Highlights: What could be better

- Clarify course scope and reading expectations — several students mention uncertainty about what constitutes the official curriculum ("pensum") and desired preparation depth.
- Balance assignment workload and pacing — a few respondents note that some exercises or lab assignments feel time-intensive or unevenly distributed.
- Enhance lecture–exercise integration — while lectures are appreciated, students suggest even tighter linkage between theory and practical sessions.



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Highlights: Group Assignment Element

- Group feedback is highly valued — students frequently describe it as useful, constructive, and motivating for improving their work.
- Collaborative dynamics are generally positive, with many appreciating how group work reinforces understanding and accountability.
- Some variation in group engagement — while most experiences are good, a few mention uneven contribution or occasional coordination challenges.



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Highlights: Open input and feedback

- Strong overall satisfaction — many spontaneous comments describe the course as fun, well-structured, and one of the best of the semester.
- Appreciation for clarity and organization — students commend the teaching structure, the GitHub materials, and the logical flow of exercises and lectures.
- Constructive suggestions — a few note a desire for more real-world data examples and slightly extended exercise time to consolidate learning.



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## 

```{r}
eval_data |>
  mk_bar_plot(x = `Quick bird's-eye view: Generally speaking, this has so far been a good course`)
```



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## 

```{r}
eval_data |>
  mk_bar_plot(x = `I would recommend this course to my fellow students / colleagues`)
```



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## 

```{r}
eval_data |>
  mk_bar_plot(x = `So far, the course content has aligned with my expectations`)
```



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## 

```{r}
eval_data |>
  mk_bar_plot(x = `So far, the course content aligns with the course learning objectives, as described in the DTU Course Base (See DTU course base description)`)
```



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## 

```{r}
time_usage_summary <- eval_data |>
  summarise(mu = mean(`In ADDITION to our 8-12 teaching sessions, how many hours are you spending OUTSIDE class per week?`),
            sigma = sd(`In ADDITION to our 8-12 teaching sessions, how many hours are you spending OUTSIDE class per week?`))
time_mu <- round(time_usage_summary$mu, 1)
time_sigma <- round(time_usage_summary$sigma, 1)
time_lower <- round(4 + time_mu - time_sigma, 0)
time_upper <- round(4 + time_mu + time_sigma, 0)

eval_data |>
  mk_bar_plot(x = `In ADDITION to our 8-12 teaching sessions, how many hours are you spending OUTSIDE class per week?`)
```

_Average: 4 + `{r} time_mu` +/- `{r} time_sigma` ~ `{r} time_lower`-`{r} time_upper`h/week, expected time usage for 5 ECTS ~9-10h/week_
