---
title: "Recap of Lab 3 Exercises"
format:
  revealjs:
    embed-resources: true
    theme: moon
    slide-number: c/t
    width: 1600
    height: 900
    mainfont: avenir
    logo: images/r4bds_logo_small.png
    footer: "R for Bio Data Science"
---



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
# Recap of Lab 3 Exercises

```{r}
#| echo: false
#| eval: true
#| message: false

library("tidyverse")
SPE_ENV <- read_tsv(file = "data/SPE_ENV.tsv")
targets <- c("Methanobacteria", "Clostridia", "Actinobacteria",
             "Sphingobacteria", "Anaerolineae")
SPE_ENV_targets <- filter(SPE_ENV, Taxa %in% targets)
```



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 1 - Recreate the following plot

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

pl <- ggplot(data = SPE_ENV,
             mapping = aes(x = OTU_Count,
                           y = Taxa,
                           fill = site)) +
  geom_boxplot()
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 2 - Recreate the following plot
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

pl <- ggplot(data = SPE_ENV,
             mapping = aes(x = site,
                           y = Temp)) +
  geom_boxplot()
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 3 - Recreate the following plot
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

pl <- ggplot(data = SPE_ENV,
             mapping = aes(x = Temp,
                           y = pH,
                           colour = site)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 4 - Recreate the following plot
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

pl <- ggplot(data = SPE_ENV,
             mapping = aes(x = Temp,
                           y = pH,
                           colour = site,
                           label = Samples)) +
  geom_label(size = 3)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 5 - Recreate the following plot
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

pl <- ggplot(data = SPE_ENV,
             mapping = aes(x = NH4,
                           fill = site)) +
  geom_density()
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 5 - Recreate the following plot - OR
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

pl <- pl +
  scale_x_log10()
```

*Note this syntax and think about what it means*

:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 6 - Recreate the following plot
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

pl <- ggplot(data = SPE_ENV,
             mapping = aes(x = CODt,
                           y = CODs,
                           colour = site)) +
  geom_point() +
  geom_smooth(method = "lm")
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 7 - Recreate the following plot
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

pl <- ggplot(data = SPE_ENV,
             mapping = aes(x = Taxa,
                           y = Samples,
                           fill = OTU_Count)) +
  geom_tile() +
  scale_fill_gradient2(midpoint = 10000,
                       low = "blue",
                       mid = "white",
                       high = "red") +
  theme_classic(base_size = 8) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))

```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true
#| fig-width: 7
#| fig-height: 7
#| fig-align: center

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 8 - Recreate the following plot
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true

library("ggridges")
pl <- ggplot(data = SPE_ENV_targets,
             mapping = aes(x = OTU_Count,
                           y = Taxa,
                           fill = Taxa)) +
  geom_density_ridges(alpha = 0.5) +
  scale_fill_viridis_d() +
  labs(x = "OTU Count",
       y = "Taxa Identified",
       title = "OTU Count Distribution for 5 Taxa Stratified on Site",
       caption = "Data from doi.org/10.1111/1751-7915.12334") +
  theme_minimal(base_family = "Avenir",
                base_size = 12) +
  theme(legend.position = "bottom") +
  facet_wrap(vars(site),
             ncol = 2)
```

:::

::: {.column width="50%"}
```{r}
#| echo: false
#| eval: true

pl
```
:::

::::



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Task 9 - <span style="color: red;">GROUP ASSIGNMENT</span>

Let us take a look at a nice example...
