---
title: "Recap of Lab 9"
author: "Søren Helweg Dam"
format:
  revealjs:
    embed-resources: true
    theme: moon
    slide-number: c/t
    width: 1600
    height: 900
    mainfont: avenir
    logo: images/r4bds_logo_small.png
    footer: "R for Bio Data Science"
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = FALSE)
```

# Exercises recap

<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Lab 9 Learning Objectives

_A student who has met the objectives of the session will be able to:_

- Prepare a simple shiny application

- Using relevant online resources to autonomously identify and obtain new and expand on existing knowledge of R


<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## How to build an R shiny app
  
- Use `golem` from the get-go
- Start with layout (UI/UX)
- Use online resources for inspiration
- Bear in mind how `declarative` programming works
  - Use the reactivity slide for inspiration
- Break your app into modules

<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Exercises

- A lot of new stuff (functions, terms, ways of programming)
- GitHub does not allow multiple people to push the same changes


<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Developing apps on the Shiny server

Shiny servers are not built for development.
</br> 

- Remember if you get a "Not Found" error, the "/shiny/" might be missing from the url.
- Your app must be in `~/ShinyApps/` to be found.
- If the app inexplicably don't work, check the log in `ShinyApps/log/` for error messages.
- If no changes appear, close the tab for a minute and try again.
- _Close your app before continuing so future changes take effect._
- Going forward, develop locally and deploy later.


<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
# Assignment recap

<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Challenges with shiny

- New challenging topic
  - More how-to examples would be great
  - Requires more than fours hours to learn
  - Complex structure and logic
  - Does it matter where to put code? [Yes, but no](https://services.healthtech.dtu.dk/shiny/sohdam/ShinyDogma/)

<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Challenges with shiny

- New challenging topic
  - More how-to examples would be great
  - Requires more than fours hours to learn
  - Complex structure and logic
  - Does it matter where to put code? [Yes, but no](https://services.healthtech.dtu.dk/shiny/sohdam/ShinyDogma/)

- Hard to collaborate
  - Git-conflict galore


<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Challenges with shiny

- New challenging topic
  - More how-to examples would be great
  - Requires more than fours hours to learn
  - Complex structure and logic
  - Does it matter where to put code? [Yes, but no](https://services.healthtech.dtu.dk/shiny/sohdam/ShinyDogma/)

- Hard to collaborate
  - Git-conflict galore

- Debugging is challenging, especially on a Shiny server
  - No error messages
  - Unclear when changes took effect


<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Pros and cons with `golem`

> _Once we got the hang of Golem, it became a valuable help for structuring and maintaining our application._

<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Pros and cons with `golem`

> _Once we got the hang of Golem, it became a valuable help for structuring and maintaining our application._

:::: {.columns}

::: {.column width="45%"}
- Encourages using modules
  - Breaks the app into small components
  - Helps testing and debugging
- Provides a streamlined and structured workflow
  - Removes repetitive coding tasks and lets you focus on app logic and structure


:::
::: {.column width="10%"}
:::
::: {.column width="45%"}



:::

::::

<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Pros and cons with `golem`

> _Once we got the hang of Golem, it became a valuable help for structuring and maintaining our application._

:::: {.columns}

::: {.column width="45%"}
- Encourages using modules
  - Breaks the app into small components
  - Helps testing and debugging
- Provides a streamlined and structured workflow
  - Removes repetitive coding tasks and lets you focus on app logic and structure


:::
::: {.column width="10%"}
:::
::: {.column width="45%"}

- Encourages using modules
  - Hard to figure out how to put components together
  - Complicated in a small app
- Steep learning curve
  - Overwhelming with so many files
  - Many functionalities to keep track of



:::

::::


<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Maintaining reproducibility

- Reproducibility is important
- Version control help managing changes
  - Show version number in the app
- Hard to keep track of decisions made by the users
  - Enforce analysis workflow



<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
## Other challenges and limitations

- Steep learning curve
- Slow performance with large datasets or many users
- Security issues
  - App server required?
- Limited graphical design?

<!--# ---------------------------------------------------------------------- -->
<!--# SLIDE ---------------------------------------------------------------- -->
<!--# ---------------------------------------------------------------------- -->
##

```{r}
#| eval: TRUE
#| echo: FALSE
knitr::include_app("https://shiny.posit.co/r/gallery/", height = 800)
```

